\name: Deterministic and Coherent Implicits

### 2018-11-19 jfp/src/Translation.tex
fix hyperref
Reviewer 1, partially addressed

The function $\ttrans$ takes 
$\name$ types $\rulet$ to System F types T: 
\begin{equation*}
\begin{array}{rcl@{\hspace{1cm}}rcl}
\ttrans[\alpha] & = & \alpha &
\ttrans[\forall \alpha. \rulet] & = & \forall \alpha. \ttrans[\rulet] \\
\ttrans[\rulet_1 \arrow \rulet_2] & = & \ttrans[\rulet_1] \arrow \ttrans[\rulet_2] &
\ttrans[\rulet_1 \iarrow \rulet_2] & = & \ttrans[\rulet_1] \arrow \ttrans[\rulet_2] %\\
\end{array}
\end{equation*}
It reveals that implicit $\name$ arrows are translated to explicit System F function arrows.

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Resolution}
The more interesting part of the translation happens when resolving queries.
Queries are translated by rule $\rref{Ty-Query}$ using the auxiliary resolution
judgment $\ivturns$:
\begin{equation*}
\dres{\tenv}{\rulet}{E}
\end{equation*}
which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.  The
translation builds a System F term as evidence for the resolution.

@@

For simple types $\type$ rule \rref{R-Simp} delegates the work of
building evidence, when a matching implicit $\rulet$ is found in the
environment, to rule \rref{L-Match}. The evidence consists of two parts:
$E$ is the evidence of matching $\type$ against $\rulet$. This match contains
placeholders $\bar{x}$ for the contexts whose resolution is postponed by rule
\rref{M-IAbs}. It falls to rule \rref{L-Match} to perform these
postponed resolutions, obtain their evidence $\bar{E}$ and fill in the
placeholders.

@@

\begin{theorem}[Type-Preserving Translation]\label{thm:type:preservation} Let $e$ be a $\name$
  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ a System F expression. If
  $\wte{\tenv}{e}{\rulet}{E}$, then $\fwte{\etrans{\tenv}}{E}{\ttrans[\rulet]}$.
\end{theorem}
Here we define the translation of the type environment form $\name$ to System F as:
\begin{equation*}
\begin{array}{rcl@{\hspace{2cm}}rcl}
\etrans{\epsilon} & = & \epsilon & \etrans{\tenv,\alpha} & = & \etrans{\tenv}, \alpha \\
\etrans{\tenv,x : \rulet} & = & \etrans{\tenv}, x : \ttrans[\rulet] &
\etrans{\tenv, \envi{\rulet}{x}} & = & \etrans{\tenv}, x : \ttrans[\rulet]
\end{array}
\end{equation*}

@@

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
\begin{theorem}[Soundness]
Figure~\ref{fig:resolution2}'s deterministic definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s ambiguous definition.
\[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]
\end{theorem}
Section~\ref{proof:det_soundness} of the Appendix points to the mechanized proof of the above
theorem.
Secondly, the deterministic resolution is effectively deterministic:
\begin{theorem}[Determinacy]\label{thm:determinacy}
%The generated evidence of resolution is uniquely determined.
Let $\tenv$ be a type environment, $\rulet$ be a type and $E_1$, $E_2$ be system F expressions.
Assume that $\unambig{}{\rulet}$ and $\forall ?\rulet_i\in\tenv$, $\unambig{}{\rulet_i}$. Then
\[ %\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad E_1 = E_2 \]
\end{theorem}
A full proof of Theorem~\ref{thm:determinacy} can be found in
Section~\ref{proof:determinism} of the Appendix.

@@

Thirdly, on top of the immediate coherence of deterministic resolution, 
an additional stability property holds.
\begin{lemma}[Stability of Resolution]\label{lem:stable:resolution}
Resolution is stable under type substitution.
\[\forall \tenv,\alpha,\tenv',\suty,\rulet, E: \enskip\
\dres{\tenv,\alpha,\tenv'}{\rulet}{E} \enskip\wedge\enskip \tenv \vdash \sigma
\enskip\Rightarrow\enskip 
\dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
\end{lemma}%
This is a key lemma to establish that static reduction of type application 
preserves typing and elaboration.
\begin{lemma}[Stability of Typing under Type Application Reduction]
\label{lem:stable:typing}
Static reduction of type application preserves typing.
\[\forall \tenv,\alpha,\sigma,\rulet, E: \enskip\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{(\Lambda \alpha.E)\,\ttrans[\suty]}
\enskip\Rightarrow\enskip 
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E[\ttrans[\suty]/\alpha]}
   \]
\end{lemma}%
This, together with another property of System~F's contextual equivalence:
\[ \forall \tenv, \alpha, E, T, T': \enskip\
     \fwte{\tenv}{(\Lambda \alpha.E)\,T}{T'} \enskip\Rightarrow\enskip 
     \fctx{\tenv}{(\Lambda \alpha.E)\,T}{E[T/\alpha]}{T'}\]
allows us to conclude the correctness of static reduction of type application.
\begin{theorem}[Correctness of Type Application Reduction]\label{lem:stable:correct}
If a type application and its reduced form elaborate to two System~F terms,
those terms are contextually equivalent.
\begin{equation*}
\forall \tenv, \alpha, e, \suty, E1, E2 : \enskip\\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{E_1} ~\wedge~
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E_2}
\enskip\Rightarrow\enskip 
\fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]}
\end{equation*}
\end{theorem}
The stability Lemmas~\ref{lem:stable:resolution} and~\ref{lem:stable:typing} are proved in
Section~\ref{proof:coherence} of the Appendix.
%-------------------------------------------------------------------------------
\subsection{Evidence Generation in the Algorithm}

@@

\begin{theorem}[Partial Correctness]
Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
Assume that $\unambig{}{\rulet}$ and also $\forall ?\rulet_i \in \tenv: \unambig{}{\rulet_i}$.
Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$.
\end{theorem}
The proof is split in two parts, one for each ``direction'' of the theorem.
Section~\ref{proof:alg_soundness} proves
soundness of the algorithm with respect to deterministic resolution, while Section~\ref{proof:alg_completeness} proves partial completeness.

@@

Now we can state the conventional type safety theorem for $\name$:
\begin{theorem}[Type Safety]
If $\wtep{\epsilon}{e}{\rulet}$, then $\mathit{eval}(e) = V$ for
some System F value $V$.
\end{theorem}
The proof follows trivially from Theorem~\ref{thm:type:preservation}.

>>>>

The function $\ttrans$ takes 
$\name$ types $\rulet$ to System F types T: 
\begin{myequation*}
\begin{array}{rcl@{\hspace{1cm}}rcl}
\ttrans[\alpha] & = & \alpha &
\ttrans[\forall \alpha. \rulet] & = & \forall \alpha. \ttrans[\rulet] \\
\ttrans[\rulet_1 \arrow \rulet_2] & = & \ttrans[\rulet_1] \arrow \ttrans[\rulet_2] &
\ttrans[\rulet_1 \iarrow \rulet_2] & = & \ttrans[\rulet_1] \arrow \ttrans[\rulet_2] %\\
\end{array}
\end{myequation*}
It reveals that implicit $\name$ arrows are translated to explicit System F function arrows.

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Resolution}
The more interesting part of the translation happens when resolving queries.
Queries are translated by rule $\rref{Ty-Query}$ using the auxiliary resolution
judgment $\ivturns$:
\begin{myequation*}
\dres{\tenv}{\rulet}{E}
\end{myequation*}
which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.  The
translation builds a System F term as evidence for the resolution.

@@

For simple types $\type$ rule \rref{R-Simp} delegates the work of
building evidence, when a matching implicit $\rulet$ is found in the
environment, to rule \rref{L-Match}. The evidence consists of two parts:
$E$ is the evidence of matching $\type$ against $\rulet$. This match contains
placeholders $\bar{x}$ for the contexts whose resolution is postponed by rule
\rref{M-IApp}. It falls to rule \rref{L-Match} to perform these
postponed resolutions, obtain their evidence $\bar{E}$ and fill in the
placeholders.

@@

\begin{theorem}[Type-Preserving Translation]\label{thm:type:preservation} Let $e$ be a $\name$
  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ a System F expression. If
  $\wte{\tenv}{e}{\rulet}{E}$, then $\fwte{\etrans{\tenv}}{E}{\ttrans[\rulet]}$.
\end{theorem}
Here we define the translation of the type environment from $\name$ to System F as:
\begin{myequation*}
\begin{array}{rcl@{\hspace{2cm}}rcl}
\etrans{\epsilon} & = & \epsilon & \etrans{\tenv,\alpha} & = & \etrans{\tenv}, \alpha \\
\etrans{\tenv,x : \rulet} & = & \etrans{\tenv}, x : \ttrans[\rulet] &
\etrans{\tenv, \envi{\rulet}{x}} & = & \etrans{\tenv}, x : \ttrans[\rulet]
\end{array}
\end{myequation*}

@@

In addition, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
Firstly, the deterministic version of resolution is a sound variation on the original ambiguous
resolution.
\begin{theorem}[Soundness]
Figure~\ref{fig:resolution2}'s deterministic definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s ambiguous definition.
\[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]
\end{theorem}
As a corollary (by use of this result with Lemma~\ref{lem:type:preservation}), the
deterministic resolution process is also type-preserving. Section~\ref{proof:det_soundness} of
the Appendix points to the mechanized proof of the above theorem.
Secondly, the deterministic resolution is effectively deterministic:
\begin{theorem}[Determinacy]\label{thm:determinacy}
%The generated evidence of resolution is uniquely determined.
Let $\tenv$ be a type environment, $\rulet$ be a type and $E_1$, $E_2$ be system F expressions.
Assume that $\unambig{}{\rulet}$ and $\forall ?\rulet_i\in\tenv$, $\unambig{}{\rulet_i}$. Then
\[ %\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad E_1 = E_2 \]
\end{theorem}
A full proof of Theorem~\ref{thm:determinacy} can be found in
Section~\ref{proof:determinism} of the Appendix.

@@

Thirdly, on top of the immediate coherence of deterministic resolution, 
an additional stability property holds.
\begin{lemma}[Stability of Resolution]\label{lem:stable:resolution}
Resolution is stable under monotype substitution.
\[\forall \tenv,\alpha,\tenv',\suty,\rulet, E: \enskip\
\dres{\tenv,\alpha,\tenv'}{\rulet}{E} \enskip\wedge\enskip \tenv \vdash \sigma
\enskip\Rightarrow\enskip 
\dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
\end{lemma}%
This is a key lemma to establish that static reduction of type application 
preserves typing and elaboration.
\begin{lemma}[Stability of Typing under Type Application Reduction]
\label{lem:stable:typing}
Static reduction of type application preserves typing.
\[\forall \tenv,\alpha,\sigma,\rulet, E: \enskip\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{(\Lambda \alpha.E)\,\ttrans[\suty]}
\enskip\Rightarrow\enskip 
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E[\ttrans[\suty]/\alpha]}
   \]
\end{lemma}%
This, together with another property of System~F's contextual equivalence:
\[ \forall \tenv, \alpha, E, T, T': \enskip\
     \fwte{\tenv}{(\Lambda \alpha.E)\,T}{T'} \enskip\Rightarrow\enskip 
     \fctx{\tenv}{(\Lambda \alpha.E)\,T}{E[T/\alpha]}{T'}\]
allows us to conclude the correctness of static reduction of type application.
\begin{theorem}[Correctness of Type Application Reduction]\label{lem:stable:correct}
If a type application and its reduced form elaborate to two System~F terms,
those terms are contextually equivalent.
\begin{myequation*}
\forall \tenv, \alpha, e, \suty, E1, E2 : \enskip\\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{E_1} ~\wedge~
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E_2}
\enskip\Rightarrow\enskip 
\fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]}
\end{myequation*}
\end{theorem}
The stability Lemmas~\ref{lem:stable:resolution} and~\ref{lem:stable:typing} are proved in
Section~\ref{proof:coherence} of the Appendix.
%-------------------------------------------------------------------------------
\subsection{Evidence Generation in the Algorithm}

@@

\begin{theorem}[Partial Correctness]
Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
Assume that $\unambig{}{\rulet}$ and also $\forall ?\rulet_i \in \tenv: \unambig{}{\rulet_i}$.
Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$.
\end{theorem}
The proof is split in two parts, one for each ``direction'' of the theorem.
Section~\ref{proof:alg_soundness} proves
soundness of the algorithm with respect to deterministic resolution, while Section~\ref{proof:alg_completeness} proves partial completeness. The latter correctness property is partial because
it does not hold without the additional termination conditions for the queried type and the
type environment, \(\tenv\).

@@

Now we can state the conventional type safety theorem for $\name$:
\begin{theorem}[Type Safety]
If $\wtep{\epsilon}{e}{\rulet}$, then $\mathit{eval}(e) = V$ for
some System F value $V$.
\end{theorem}
The proof follows from Theorem~\ref{thm:type:preservation} and System F's well-known
normalization property.
